---
title: "goldval"
output: github_document
---

`goldval` is an early-stage R workflow for clinical prediction model validation
when a proxy outcome is observed for everyone and a gold-standard outcome is
available only for a reviewed subset.

This package is positioned as biomedical software/workflow infrastructure, not
as a claim of new AUC, Brier, or calibration-slope estimators.

## Problem

Clinical prediction models are often validated using convenient outcomes from
ICD codes, EHR phenotypes, claims, or registries. These proxy outcomes can be
wrong. A smaller chart-reviewed subset may contain the gold-standard outcome,
but review is often selective: proxy-positive or high-risk patients are more
likely to be reviewed.

```text
             Full validation cohort
        +-----------------------------+
        | prediction p                |
        | proxy outcome Y*            |
        +-----------------------------+
                      |
            selective chart review
                      |
                      v
             +----------------+
             | gold outcome Y |
             +----------------+
```

## Quick Start

```{r eval=FALSE}
library(goldval)

x <- goldval(
  data = validation_data,
  pred = "predicted_risk",
  proxy_outcome = "ehr_outcome",
  gold_outcome = "chart_review_outcome",
  verified = "reviewed"
)

diagnose_goldval(x)
```

`goldval()` also accepts direct vectors:

```{r eval=FALSE}
x <- goldval(
  pred = validation_data$predicted_risk,
  proxy_outcome = validation_data$ehr_outcome,
  gold_outcome = validation_data$chart_review_outcome,
  verified = validation_data$reviewed
)
```

## Current MVP Scope

Implemented:

- `goldval()`: validated object construction;
- `diagnose_goldval()`: selective verification and positivity diagnostics.

Next functions:

- `performance()`;
- `calibration()`.

Explicit non-goals:

- no novel AUC estimator claim;
- no novel Brier estimator claim;
- no novel calibration-slope/intercept estimator claim;
- no DCA, NRI/IDI, survival, fairness, SHAP, or simultaneous bands in v0.1.
