---
title: "goldval workflow with synthetic data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{goldval workflow with synthetic data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(goldval)
```

## The problem

Clinical prediction models are often evaluated in data where the model's
predicted risk is available for every patient, but the best outcome label is
not. A common structure is:

- a fixed predicted risk for every patient;
- an error-prone proxy outcome for every patient;
- a gold-standard outcome only for a reviewed subset;
- a review process that may depend on observed patient information.

The goal of `goldval` is to organize this validation workflow. It does not
claim new estimator theory.

## Generate a synthetic example

This vignette uses fully synthetic data. The true outcome is kept only so we can
compute an oracle simulation reference; real applications do not have this
column for the full cohort.

```{r synthetic-data}
set.seed(20260808)

n <- 1800L
x1 <- rnorm(n)
x2 <- rbinom(n, 1, 0.45)

true_risk <- plogis(-1.35 + 1.15 * x1 + 0.65 * x2)
true_outcome <- rbinom(n, 1, true_risk)

pred <- plogis(-1.20 + 0.95 * x1 + 0.55 * x2 + rnorm(n, sd = 0.25))

proxy <- ifelse(
  true_outcome == 1L,
  rbinom(n, 1, 0.82),
  rbinom(n, 1, 0.10)
)

lp <- as.numeric(scale(qlogis(pmin(pmax(pred, 1e-6), 1 - 1e-6))))
eta_no_intercept <- log(4) * proxy + 0.75 * lp
target_gold_n <- 250L
target_rate <- target_gold_n / n
review_intercept <- uniroot(
  function(a) mean(plogis(a + eta_no_intercept)) - target_rate,
  interval = c(-20, 20)
)$root
review_probability <- plogis(review_intercept + eta_no_intercept)
verified <- rbinom(n, 1, review_probability)

gold <- ifelse(verified == 1L, true_outcome, NA)

dat <- data.frame(
  pred = pred,
  proxy = proxy,
  gold = gold,
  verified = verified,
  true_outcome = true_outcome
)

mean(dat$verified)
sum(dat$verified)
```

## Construct a goldval object

```{r goldval-object}
g <- goldval(
  pred = dat$pred,
  proxy_outcome = dat$proxy,
  gold_outcome = dat$gold,
  verified = dat$verified
)

g
```

## Diagnose verification

`diagnose_goldval()` summarizes the reviewed subset, proxy enrichment,
positivity, and nuisance-model status.

```{r diagnose}
diag <- diagnose_goldval(g)
diag
```

The review process is selective in this simulation: patients with higher
predicted risk are more likely to receive gold-standard review.

```{r review-by-risk}
risk_group <- cut(dat$pred, breaks = quantile(dat$pred, probs = seq(0, 1, 0.1)), include.lowest = TRUE)
review_by_risk <- aggregate(dat$verified, by = list(risk_decile = risk_group), FUN = mean)

plot(
  seq_len(nrow(review_by_risk)),
  review_by_risk$x,
  type = "b",
  xlab = "Predicted-risk decile",
  ylab = "Gold-standard review fraction",
  ylim = c(0, max(review_by_risk$x) * 1.15)
)
```

## Performance

`performance()` compares validation estimates using the proxy outcome, the
reviewed gold subset, and corrected analyses.

```{r performance}
perf <- performance(
  g,
  method = c("naive", "gold_only", "OR", "AIPW"),
  metrics = c("auc", "brier")
)

perf
```

Because this is a simulation, we can compute an oracle reference using the full
true outcome. This reference is unavailable in real validation studies.

```{r oracle-reference}
oracle <- goldval(
  pred = dat$pred,
  proxy_outcome = dat$true_outcome,
  gold_outcome = ifelse(dat$verified == 1L, dat$true_outcome, NA),
  verified = dat$verified
)

oracle_perf <- performance(oracle, method = "naive", metrics = c("auc", "brier"))
oracle_perf$estimates$method <- "oracle_simulation_reference"
oracle_perf$estimates
```

## Calibration

`calibration()` returns weak calibration parameters and flexible calibration
curves. The curve plot is often the most useful first look at how proxy,
gold-only, and corrected validation differ.

```{r calibration}
cal <- calibration(
  g,
  method = c("naive", "gold_only", "OR", "AIPW"),
  type = c("weak", "curve")
)

cal
```

```{r calibration-plot}
plot(cal)
```

## Scalar uncertainty

`bootstrap_goldval()` computes percentile bootstrap intervals for scalar
estimates. This example uses a small number of bootstrap replicates only to keep
the vignette fast; substantive analyses should use a larger value chosen for the
intended application.

```{r bootstrap}
boot <- bootstrap_goldval(
  g,
  method = "AIPW",
  metrics = c("auc", "brier"),
  B = 50,
  seed = 20260808
)

boot
```

The bootstrap object also stores replicate-level diagnostics and
method/metric-level failure summaries.

`interval_status = "ok"` indicates that enough bootstrap replicates were
successfully estimated; it does not by itself establish adequate positivity or
nuisance-model stability. Review `diagnose_goldval()` and bootstrap diagnostics
alongside the interval.

```{r bootstrap-diagnostics}
boot$failure_summary
head(boot$replicate_diagnostics)
```

## Limitations

This vignette demonstrates the user-facing workflow rather than new statistical
theory. Current boundaries are:

- flexible calibration curve confidence intervals are out of scope;
- percentile uncertainty has been validated only in selected AIPW simulation
  scenarios;
- extreme selective verification with small gold-standard samples may yield
  undercoverage for some metrics, especially Brier score or weak calibration
  slope;
- nuisance-model warnings should be reviewed rather than ignored;
- the software does not establish novel estimator theory.
